---
title: "Reproducible Research: Peer Assessment 1"
output: html_document
html_document: PA1_template.html
keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE)
```

```{r includ_library, include=FALSE}
library(reshape2)
library(ggplot2)
library(lubridate)
```

## Loading and preprocessing the data


We set the working directory followed by downloading the dataset from its url and unzipping the file to "step_data.csv". The data comes from Roger Pengs [github account](https://github.com/rdpeng/RepData_PeerAssessment1).

```{r load data}
setwd("C:\\myCode-R")
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
dest.file <- "step_data.zip"
download.file(url, dest.file)
unzip(dest.file)
activity <- read.csv("activity.csv", sep = ",")
```

The variable names and the structure of the file are -
```{r activity details}
names(activity)

str(activity)

# Removing NA rows from given data set
head(activity[which(!is.na(activity$steps)), ])
```

Now data is ready for further analysis.

## What is mean total number of steps taken per day?

Group the number of steps by date and intervals. Find the total number of steps per day over all days. Note that some of the days such as 2012-10-01 have no steps data. Remove such rows for this part.

```{r steps mean}

activity.melt <- melt(activity[which(!is.na(activity$steps)), ], id.vars = c("date", "interval"))
head(activity.melt)

total.steps <- dcast(activity.melt, date ~ variable, sum)
head(total.steps)
```

Then we can find the mean of 'total number of steps per day'.
```{r steps summary}
steps.summary <- summary(total.steps$steps)
names(steps.summary)
step.mean <- steps.summary["Mean"]
```

Histogram of the total number of steps taken each day -
```{r total step per day histogram}

hist(total.steps$steps, 
     main = "Histogram of total steps taken per day",
     xlab = "Total steps per day", 
     ylab = "Number of days", 
     breaks = 10, 
     col = "steel blue")

abline(v = mean(total.steps$steps), lty = 1, lwd = 2, col = "red")

abline(v = median(total.steps$steps), lty = 2, lwd = 2, col = "black")

legend(x = "topright", 
       c("Mean", "Median"), 
       col = c("red", "black"),
       lty = c(1, 2), 
       lwd = c(2, 2))
```

Equivalent ggplot looks like -
```{r ggplot histogram}

ggplot(total.steps, aes(steps)) + geom_histogram(bins = 10)
```

Here's a plot showing the trend in total number of steps taken per day over two months -
```{r total steps trend}

total.steps$date <- as.Date(total.steps$date)

ggplot(total.steps, 
       aes(date, steps)) + 
  geom_line() + 
  scale_x_date(date_labels = "%b %d") + 
  ylab("Total number of steps")
```

**Conclusion:** 'r step.mean' : mean of steps taken per day.

## What is the average daily activity pattern?

- Calculate average steps for each interval for all days.
- Plot the Average Number Steps per Day by Interval.
- Find interval with most average steps.
- Makeing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken averaged across all days.


```{r find max steps}
steps.mean.interval <- dcast(activity.melt, interval ~ variable, mean, na.rm = TRUE)
head(steps.mean.interval)

plot(steps.mean.interval$interval, 
     steps.mean.interval$steps, ty = "l", 
     xlab = "Interval (time)",
     ylab = "Average steps",
     main = "Average steps taken over all days vs \n time interval")

# Finding maximum number of steps.
max.steps.interval <- steps.mean.interval$interval[which.max(steps.mean.interval$steps)]
max.steps.interval
```

**Conclusion:** The 5-minute interval, on average across all the days in the data set, containing the maximum number of steps is 'r max.steps.interval'.

## Imputing missing values

Lets find missing values and get a sense of that. We can replace the missing data for a day by the time average over all other days. 

```{r impute missing data}
## Impute missing values
activity2 <- split(activity, activity$interval)

activity2 <- lapply(activity2, 
                    function(x) { 
                      x$steps[which(is.na(x$steps))] <- mean(x$steps, na.rm = TRUE) 
                      return(x) 
                      }
                    )

activity2 <- do.call("rbind", activity2)
row.names(activity2) <- NULL

activity2 <- split(activity2, activity2$date)

activity2 <- lapply(activity2, 
                      function(x) {
                        x$steps[which(is.na(x$steps))] <- mean(x$steps, na.rm = TRUE)
                        return(x)
                        }
                      )

activity2 <- do.call("rbind", activity2)
row.names(activity2) <- NULL
head(activity2)
```

Assuming that the time intervals form a disjoint partitioning of 24 hrs, i.e. 1 day is found to be erroneous. The time interval for each day corresponds to approximately 40 hours, which refutes the intervals being disjoint.

```{r steps mean2}
activity.melt.2 <- melt(activity2, id.vars = c("date", "interval"))
total.steps.2 <- dcast(activity.melt.2, date ~ variable, sum, na.rm = TRUE)
head(total.steps.2)
```

Histogram of the total number of steps taken each day with the imputed missing values.

```{r total step per day histogram2}
hist(total.steps.2$steps, 
     main = "Histogram of total steps taken per day", 
     xlab = "Total steps per day", 
     ylab = "Number of days", 
     breaks = 10, 
     col = "steel blue")

abline(v = mean(total.steps.2$steps), lty = 1, lwd = 2, col = "red")
abline(v = median(total.steps.2$steps), lty = 2, lwd = 2, col = "black")
legend(x = "topright", 
       c("Mean", "Median"), 
       col = c("red", "black"), 
       lty = c(2, 1), 
       lwd = c(2, 2))
```

Number of rows with NA values -

```{r NA row num}
sum(is.na(activity$steps))

# Percentage of rows with missing values
sum(is.na(activity$steps))*100/nrow(activity) 
```

## Are there differences in activity patterns between weekdays and weekends?

Create a new column describing if the date is a weekday or weekend.

```{r weekday status}

weekends <- which(weekdays(as.Date(activity2$date)) == "Saturday" |
              weekdays(as.Date(activity2$date)) == "Sunday")
weekdays <- which(weekdays(as.Date(activity2$date)) != "Saturday" &
              weekdays(as.Date(activity2$date)) != "Sunday")

temp.day <- c(rep("a", length(activity2)))
temp.day[weekends] <- "weekend" # Assign Weekend day name
temp.day[weekdays] <- "weekday" # Assign Weekday day name
length(temp.day) # Find a length
names(temp.day) <- "day" # Assign name

activity2 <- cbind(activity2, temp.day)
names(activity2)[4] <- "day" # Assign column heading
head(activity2)
```

Steps taken over each interval averaged across weekday days and weekend days.

```{r}
day.split.activity2 <- split(activity2, activity2$day)
steps.mean.interval <- lapply(day.split.activity2, 
                              function(x) {
                                temp <- aggregate(x$steps, list(x$interval), mean)
                                names(temp) <- c("interval", "steps")
                                return(temp)
                                }
                              )

## Unsplit stepsmean_interval
steps.mean.interval <- do.call("rbind", steps.mean.interval)
weekdays <- grep("weekday" ,row.names(steps.mean.interval))
weekends <- grep("weekend" ,row.names(steps.mean.interval))
temp <- c(rep("a", length(steps.mean.interval$steps)))
temp[weekdays] <- "weekdays"
temp[weekends] <- "weekends"
steps.mean.interval <- cbind(steps.mean.interval, temp)
row.names(steps.mean.interval) <- NULL
names(steps.mean.interval)[3] <- "day"
head(steps.mean.interval)
tail(steps.mean.interval)

ggplot(steps.mean.interval, aes(interval, steps)) + geom_line() + facet_grid(day ~ .) 
```

The mean number of steps taken over the weekdays and weekends.

```{r weekday/end mean steps}
steps.data.melt <- melt(steps.mean.interval, id.vars = c("interval", "day"))
day.step.avg <- dcast(steps.data.melt, day ~ variable, mean) # Average steps
```

**Conclusion:** We have observed that 'r day.step.avg[1,1]' step average is 'r day.step.avg[1,2]', however 'r day.step.avg[2,1]' step average is 'r day.step.avg[2,2]'.
